---
title: "class13"
author: "Ann Truong (PID: A16028103)"
format: pdf
---

## Differential Section Analysis 

```{r}
library(DESeq2)
```

Let's load our data and take a look at it.
```{r}
# Import metadata and take a peak
colData = read.csv("GSE37704_metadata.csv", row.names=1)
head(colData)
```

```{r}
# Import countdata
countData = read.csv("GSE37704_featurecounts.csv", row.names=1)
head(countData)
```

```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[,-1])
head(countData)
```

```{r}
# Filter count data where you have 0 read count across all samples.
zerocounts <- rowSums(countData) == 0
head(zerocounts)
```

```{r}
newcounts <- countData[ !zerocounts, ]
head(newcounts)
```

```{r}
nrow(newcounts)
```

**Running DESeq2**

Nice now lets setup the DESeqDataSet object required for the DESeq() function and then run the DESeq pipeline.

```{r}
dds = DESeqDataSetFromMatrix(countData=newcounts,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
## estimating size factors
## estimating dispersions
## gene-wise dispersion estimates
## mean-dispersion relationship
## final dispersion estimates
## fitting model and testing

dds
```

```{r}
colData = read.csv("GSE37704_metadata.csv", row.names=1)
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))

summary(res)
```

**Volcano plot**
A plot of log2 fold change vs -log adjusted p-value.

```{r}
plot( res$log2FoldChange, -log(res$padj) )

# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
mycols[ res$padj<0.05 & abs(res$log2FoldChange) > 2] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

**Adding gene annotation**

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)
```

```{r}
res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```

## Pathway analysis 

```{r}
library(pathview)
```

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

The main gage() function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

Now, let’s run the gage pathway analysis.

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
```

```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```

Now, let's try out the pathview() function from the pathview package to make a pathway plot with our RNA-Seq expression results shown in color.

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

Now, let's process our results a bit more to automagicaly pull out the top 5 upregulated pathways, then further process that just to get the pathway IDs needed by the pathview() function.

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

Finally, lets pass these IDs in keggresids to the pathview() function to draw plots for all the top 5 pathways.

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```
Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?
Yes, we can just manipulate the code chunk from `rownames(keggres$greater)` to `rownames(keggres$lesser)`.