---
title: "class10"
author: "Ann (PID: A16028103)"
format: pdf
---

# Comparative analysis of ADK

We will get an example ADK sequence from the database. Then, we will use this to find all ADK structures in PDB. 

```{r}
# Install packages in the R console NOT your Rmd/Quarto file

# install.packages("bio3d")
# install.packages("devtools")
# install.packages("BiocManager")

BiocManager::install("msa")
devtools::install_bitbucket("Grantlab/bio3d-view")
```


```{r}
library("bio3d")
library("devtools")
library("BiocManager")
library("msa")
```

**Q10. Which of the packages above is found only on BioConductor and not CRAN?** msa

**Q11. Which of the above packages is not found on BioConductor or CRAN?** 

**Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?**

```{r}
aa <- get.seq("1ake_A")
aa
```

We can now run BLAST with this sequence. 

```{r}
b <- blast.pdb(aa)
```


```{r}
# Plot a summary of search results
hits <- plot(b)

# Let's see what's in our `hits` object
# hits$pdb.id)

hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```

```{r}
# Download related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)

# Align related PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

Now that we have aligned & superposed structures, we can perform all sorts of analysis on them. Let's do PCA...

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

Results of PCA on ADK structures. Each dot represents 1 PDB structure.

We can cluster the structures by RMSD (or any other method).

```{r}
# Measuring / calculating distance
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col=grps)
```

We want to visualize more major structural variations.

```{r}
# Visualize first principal component, open in Molestar
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```



